Understanding Subsidy Allocation in Toronto’s Child Care Centres: Insights from a Logistic Regression Analysis

### 2024-12-11 paper/paper.qmd
updated citation
updated figure caption and beautified graphs and tables

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
data <- read_parquet(here("data","02-analysis_data","analysis_data.parquet"))
```

@@

# Create a GT table with the first 10 rows
data %>%
  head(10) %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Overview of Cleaned Data",
    subtitle = "Displaying the first 10 rows"
  )
```

@@

```{r}
#| label: fig-subsidy
#| fig-cap: "Distribution of Subsidy Allocation: Disparities Exist"
#| echo: false
#| message: false
#| warning: false

@@

# Create the bar plot with percentage labels
plot1 <- ggplot(data_summary, aes(x = factor(subsidy, labels = c("No", "Yes")), y = count)) +
  geom_bar(stat = "identity", fill = c("skyblue", "coral"), alpha = 0.8) +
  geom_text(aes(label = paste0(percentage, "%")), vjust = -0.5, size = 3) +  # Add percentage labels
  labs(
    title = "Distribution of Subsidy",
    x = "Subsidy",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 9),  # Reduce title size
    axis.title = element_text(size = 10),  # Reduce axis title size
    axis.text = element_text(size = 8)    # Reduce axis text size
  )

@@

- `pROC` [@citeproc]: Facilitated the computation and visualization of Receiver Operating Characteristic (ROC) curves and Area Under the Curve (AUC) metrics, enabling robust evaluation of the model's classification performance.

@@

```{r}
#| label: fig-relationbldgtype
#| fig-cap: "Subsidy Allocation by Building Type: Higher Subsidy Proportions in Public and Purpose-Built Facilities"
#| echo: false
#| message: false
#| warning: false
# Update the plot to use descriptive labels for the subsidy legend
ggplot(data, aes(x = bldg_type, fill = as.factor(subsidy))) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(
    values = c("blue", "red"),
    labels = c("No", "Yes")  # Replace 0/1 with No/Yes
  ) +
  labs(
    x = "Building Type",
    y = "Proportion",
    fill = "Subsidy Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate x-axis labels for readability
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )
```

@@

ggplot(data, aes(x = AUSPICE, fill = as.factor(subsidy))) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(
    values = c("blue", "red"),
    labels = c("No", "Yes")  # Change legend labels to No/Yes
  ) +
  labs(
    x = "Auspice",
    y = "Proportion",
    fill = "Subsidy Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )
```

@@

# Plot the heatmap without axis labels
ggcorrplot(cor_matrix, lab = TRUE) +
  scale_x_discrete(labels = c("Total Space", "Subsidy", "CWELCC Participation")) +  # Update x-axis labels
  scale_y_discrete(labels = c("Total Space", "Subsidy", "CWELCC Participation")) +  # Update y-axis labels
  labs(
    title = "Correlation Heatmap of Key Variables",
    caption = "Key Variables: Total Space, Subsidy, and CWELCC Participation"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank()   # Remove y-axis title
  )

```

@@

   - Represents the baseline log-odds of subsidy allocation when all predictors are set to their reference categories or zero values.
   - It provides the starting point for predicting subsidy probabilities.

@@

# Load model 
model <- readRDS(file = here("models","final_model.rds"))

@@

The regression model reveals several key observations regarding the predictors and overall goodness of fit. The intercept, estimated at -4.818, represents the baseline log-odds of the outcome when all predictors are at their reference or zero level.  While not directly interpretable, it serves as a baseline reference for the model. Among the predictors, the operating auspice: non-profit Agency category significantly increases the log-odds of the outcome, with an estimate of 2.937 and a small standard error of 0.192, indicating a strong and reliable positive effect. However, the operating auspice: other category shows an unusually large coefficient (18.680) paired with a very high standard error (631.207), suggesting instability. The CWELCC participation variable also exhibits a strong and reliable positive effect, with an estimate of 3.101 and a standard error of 0.346, indicating that being flagged as CWELCC significantly increases the log-odds of the outcome.  Additionally, the total space variable has a small but statistically significant effect, with an estimate of 0.013 and a low standard error of 0.003, reflecting robustness. In terms of model fit, the dataset includes 750 observations, and the model's AIC (760.8) and BIC (785.7) suggest a reasonable balance between goodness of fit and complexity, as lower values are generally preferred [@burnhammodel].  The log-likelihood of -375.414 also supports an adequate fit, with a higher (less negative) value indicating better alignment between the model and the data.  Finally, the RMSE of 0.32 reflects the model's predictive accuracy, with a low value indicating that the model's predictions closely match the observed data.  Overall, the model performs well but may require refinement, particularly in addressing instability in the AUSPICE: Other variable.
@fig-cp further visualizes the significance of predictors.

@@

# Plot with enhancements
ggplot(tidy_model, aes(x = estimate, y = reorder(term, abs(estimate)), color = significant)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "black")) +
  labs(
    title = "Coefficient Plot",
    x = "Coefficient Estimate",
    y = "Predictor Variables",
    color = "Significance"
  ) +
  theme_minimal()
```

@@

# ROC Curve and AUC
roc_curve <- roc(data$subsidy, 
                 predict(model, 
                         newdata = data, type = "response"))
plot(roc_curve, main = "ROC Curve", print.auc=T)
```

@@

```{r}
#| label: fig-pvo
#| fig-cap: " Predicted vs. Observed Probability Plot: Model Calibration and Accuracy"
#| echo: false
#| message: false
#| warning: false
# Generate predicted probabilities for the test dataset
data$predicted_prob <- predict(model, newdata = data, type = "response")

>>>>

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
data <- read_parquet(here("data", "02-analysis_data", "analysis_data.parquet"))
```

@@

# Create a GT table with the first 10 rows
data %>%
  rename(
    Ward = ward,
    `Operating Auspice` = AUSPICE,
    `Building Type` = bldg_type,
    `CWELCC Participation` = cwelcc_flag,
    `Total Space` = TOTSPACE,
    Subsidy = subsidy
  ) %>%
  head(10) %>%
  gt::gt()
```

@@

```{r}
#| label: fig-subsidy
#| fig-cap: "Distribution of subsidy allocation among licensed child care centres, indicating that 67.3% of centres receive subsidies while 32.7% do not."
#| echo: false
#| message: false
#| warning: false

@@

# Create the bar plot with percentage labels
plot1 <- ggplot(data_summary, aes(x = factor(subsidy, labels = c("No", "Yes")), y = count)) +
  geom_bar(stat = "identity", fill = c("skyblue", "coral"), alpha = 0.8) +
  geom_text(aes(label = paste0(percentage, "%")), vjust = -0.5, size = 3) + # Add percentage labels
  labs(
    x = "Subsidy",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 9), # Reduce title size
    axis.title = element_text(size = 10), # Reduce axis title size
    axis.text = element_text(size = 8) # Reduce axis text size
  )

@@

- `pROC` [@citepROC]: Facilitated the computation and visualization of Receiver Operating Characteristic (ROC) curves and Area Under the Curve (AUC) metrics, enabling robust evaluation of the model's classification performance.

@@

```{r}
#| label: fig-relationbldgtype
#| fig-cap: "Subsidy Allocation by Building Type: Higher Subsidy Proportions in Public and Purpose-Built Facilities"
#| echo: false
#| message: false
#| warning: false
# Update the plot to use descriptive labels for the subsidy legend
ggplot(data, aes(x = bldg_type, fill = as.factor(subsidy))) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(
    values = c("blue", "red"),
    labels = c("No", "Yes") # Replace 0/1 with No/Yes
  ) +
  labs(
    x = "Building Type",
    y = "Proportion",
    fill = "Subsidy"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1), # Rotate x-axis labels for readability
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )
```

@@

ggplot(data, aes(x = AUSPICE, fill = as.factor(subsidy))) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(
    values = c("blue", "red"),
    labels = c("No", "Yes") # Change legend labels to No/Yes
  ) +
  labs(
    x = "Auspice",
    y = "Proportion",
    fill = "Subsidy"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )
```

@@

# Plot the heatmap without axis labels
ggcorrplot(cor_matrix, lab = TRUE) +
  scale_x_discrete(labels = c("Total Space", "Subsidy", "CWELCC Participation")) + # Update x-axis labels
  scale_y_discrete(labels = c("Total Space", "Subsidy", "CWELCC Participation")) + # Update y-axis labels
  theme_minimal() +
  theme(
    axis.title.x = element_blank(), # Remove x-axis title
    axis.title.y = element_blank() # Remove y-axis title
  )
```

@@

   - Represents the baseline log-odds of subsidy allocation when all predictors are set to their reference categories or zero values.

   - It provides the starting point for predicting subsidy probabilities.

@@

# Load model
model <- readRDS(file = here("models", "final_model.rds"))

@@

The regression model reveals several key observations regarding the predictors and overall goodness of fit. The intercept, estimated at -4.818, represents the baseline log-odds of the outcome when all predictors are at their reference or zero level.  While not directly interpretable, it serves as a baseline reference for the model. Among the predictors, the operating auspice: non-profit Agency category significantly increases the log-odds of the outcome, with an estimate of 2.937 and a small standard error of 0.192, indicating a strong and reliable positive effect. However, the operating auspice: other category shows an unusually large coefficient (18.680) paired with a very high standard error (631.207), suggesting instability. The CWELCC participation variable also exhibits a strong and reliable positive effect, with an estimate of 3.101 and a standard error of 0.346, indicating that being flagged as CWELCC significantly increases the log-odds of the outcome.  Additionally, the total space variable has a small but statistically significant effect, with an estimate of 0.013 and a low standard error of 0.003, reflecting robustness. In terms of model fit, the dataset includes 750 observations, and the model's AIC (760.8) and BIC (785.7) suggest a reasonable balance between goodness of fit and complexity, as lower values are generally preferred [@burnhammodel].  The log-likelihood of -375.414 also supports an adequate fit, with a higher (less negative) value indicating better alignment between the model and the data.  Finally, the RMSE of 0.32 reflects the model's predictive accuracy, with a low value indicating that the model's predictions closely match the observed data.  Overall, the model performs well but may require refinement, particularly in addressing instability in the AUSPICE: Other variable.

@fig-cp further visualizes the significance of predictors.

@@

# Plot with enhancements
ggplot(tidy_model, aes(x = estimate, y = reorder(term, abs(estimate)), color = significant)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "black")) +
  labs(
    x = "Coefficient Estimate",
    y = "Predictor Variables",
    color = "Significance"
  ) +
  theme_minimal()
```

@@

# ROC Curve and AUC
roc_curve <- roc(
  data$subsidy,
  predict(model,
    newdata = data %>%
      rename(
        Ward = ward,
        `Operating_Auspice` = AUSPICE,
        `Building_Type` = bldg_type,
        `CWELCC_Participation` = cwelcc_flag,
        `Total_Space` = TOTSPACE
      ), type = "response"
  )
)
plot(roc_curve, main = "ROC Curve", print.auc = T)
```

@@

```{r}
#| label: fig-pvo
#| fig-cap: " Predicted vs. Observed Probability Plot: Model Calibration and Accuracy"
#| echo: false
#| message: false
#| warning: false
# Generate predicted probabilities for the test dataset
data$predicted_prob <- predict(model, newdata = data %>%
  rename(
    Ward = ward,
    `Operating_Auspice` = AUSPICE,
    `Building_Type` = bldg_type,
    `CWELCC_Participation` = cwelcc_flag,
    `Total_Space` = TOTSPACE
  ), type = "response")
